### 2022-09-27 main/chapter_context.tex
write some on dark matter

\subsection{Opportunities for future Standard Models}
Problems with the Standard Model mean it needs replacing with supersymmetry
--- is currently a false statement.
Such an argument could be tempting to motivate this thesis, but the
Standard Model is actually good in its domain.
This discussion therefore looks towards where an altered Standard Model could
go to improve what it already does, or to expand its generality.

@@

\paragraph{Neutrino masses.}
Neutrinos could be Dirac fermions with Yukawa masses and boring right-chiral
fields.
Or they could also have Majorana mass terms, for which the seesaw mechanism
can simultaneously generate the observed light neutrinos and heavy `sterile'
neutrinos.
Or their masses could come from other operators without right-handed
fields~\cite{
thomson2013modern,
chala2021neutrino,
wells2020discovery
}.
This riddle is largely unknown, and new results from theory or experiment
that pinpoint a correct model could improve the Standard Model in the neutrino
domain.

\paragraph{Gravitation and cosmology.}

\TODO{gravity}

\TODO{dark matter: rotation curves~\cite{
begeman1991rotation,
garrett2010dark
}}

\TODO{
sterile neutrinos~\cite{boyarsky2019sterile},
primordial black holes~\cite{bernard2019primordial},
WIMPs~\cite{jungman1996wimp},
axions~\cite{kim2008axions},
dust?~\cite{garrett2010dark}
}
\TODO{combinations!}

\TODO{
dark matter,
cosmological constant~\cite{riess1998observational},
inflation~\cite{bezrukov2008higgs},
baryogenesis (matter-antimatter asymmetry)
}
\TODO{note \lcdm}

>>>>

\subsection{Opportunities for future Standard Models}
Problems with the Standard Model mean it needs replacing with supersymmetry
--- is a false statement.
Such an argument could motivate this thesis, but the Standard Model is actually
good in its domain.
This discussion therefore looks towards where an altered Standard Model could
go to improve its current abilities or to expand its generality.

@@

\paragraph{Neutrino masses.}
Neutrinos could be Dirac fermions with Yukawa masses and boring right-chiral
fields.
Or they could also have Majorana mass terms, for which the seesaw mechanism
can simultaneously generate the observed light neutrinos and heavy `sterile'
neutrinos.
Or their masses could come from other operators without right-handed
fields~\cite{
thomson2013modern,
chala2021neutrino,
wells2020discovery
}.
The answer is uncertain, and new results from theory or experiment that
pinpoint a correct model could improve the Standard Model in the neutrino
domain.

\paragraph{Gravitation and cosmology.}
Gravity does not feature in the standard model, nor does it measurably impact
particle physics experiments.
Although gravitation is in another domain, the particle content of the
Standard Model does impact cosmology.

The universe contains dark matter which interacts with gravity but not much
with other forces; it is observed through galactic rotation curves,
gravitational lensing, and cosmic microwave background data,
but its composition is uncertain~\cite{
begeman1991rotation,
garrett2010dark,
planck2020results
}.
Dark matter candidates include new particles such as
sterile neutrinos~\cite{boyarsky2019sterile},
WIMPs (Weakly Interacting Massive Particles)~\cite{jungman1996wimp},
or
axions~\cite{kim2008axions},
but also macroscopic objects such as
primordial black holes~\cite{bernard2019primordial}.
Dark matter may well comprise various of different components, and it would
be a success for future particle physics theories to pinpoint necessary dark
matter particles.

\TODO{
Three more mysteries:
cosmological constant~\cite{riess1998observational},
inflation~\cite{bezrukov2008higgs},
baryogenesis (matter-antimatter asymmetry)
}
